Supermartingales and Termination Analysis of Statistical PCF

### 2021-04-29 LICS21/antitone.tex
Remove appendices
Cut down the length some more

\begin{example}[Random walk]
\label{ex:ac-ranking}
\begin{compactenum}[(i)]
\item One-dimensional (1D) biased (towards 0) random walk
\[
M_1 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 1)}\big) \, 10
\]

\item 1D unbiased random walk
\[
M_2 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (away from 0) random walk
\[
M_3 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
\]
%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
\end{compactenum}
\end{example}

\medskip

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking \Cref{thm:confluent ranking}, as can the others in this section with similar issues. See \cref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.

>>>>

\begin{example}[Random walks]
\label{ex:ac-ranking}
\begin{align*}
M_1 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 1)}\big) \, 10 \\
M_2 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10 \\
M_3 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
\end{align*}
\end{example}

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking \Cref{thm:confluent ranking}, as can the others in this section with similar issues. See \cref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
Now
\[
g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2  >  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\]
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in \cite{arxiv}.

### 2021-04-29 LICS21/confluent.tex
Cut down the length some more

In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, we first prove the relation between these different strategies, providing a confluent version of the semantics. It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies. Positions are defined as a way of addressing sample statements within a program independently of the reduction order. Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in. The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term. In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices. The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.

Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
to which both $B_1$ and $B_2$ reduce,
so the reduction order mostly doesn't matter. 
In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction. 
There are, however, some restricted variations on the reduction order that do not have this problem.

@@

Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction \emph{takes place at $\alpha$}. Each subterm can be a redex in at most one way, but there can be multiple redexes at different positions.
\end{definition}

The argument of a $\beta$ redex and the body of a $\tY$ redex may be duplicated by those reductions. It is therefore these cases that need to be handled carefully to avoid duplicating samples at the wrong time. In both cases, the potentially duplicated part must already be a value, which excludes terms like $\tsample$ or $\tsample + \underline 1$, which should be evaluated before being duplicated. In the other direction, if a $\tsample$ occurs inside of a $\lambda$, it may need to be duplicated before being evaluated, which is why a $\tsample$ reduction isn't allowed inside a $\lambda$ inside a $\tY$ or the right side of an application. These restrictions are in some cases unnecessarily strict, for example, in $(\lambda x. x) ((\lambda y. \tsample) \underline 0)$, it would be fine to evaluate the sample first, but they are at least sufficient to ensure confluence in terms of the distribution of results. Getting individual traces to behave correctly will take more work though.

>>>>

In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual strategy, we prove the relation between these different strategies, providing a confluent version of the semantics. It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies. Positions are defined as a way of addressing subterms independently of the reduction order. Next, we present a nondeterministic version of the reduction relation, that allows a choice of what order to perform reductions in. The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term. In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices. The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.

Non-probabilistic lambda calculi are generally confluent, i.e.~if $A \to B_1, B_2$, then $\exists C.\ B_1, B_2 \to^* C$,
so the reduction order mostly doesn't matter. 
In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction. 
There are, however, some restricted variations on the reduction order that do not have this problem.

@@

Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$} \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction \emph{takes place at $\alpha$}. Each subterm can be a redex in at most one way, but there can be multiple redexes at different positions.
\end{definition}

### 2021-04-29 LICS21/intro.tex
Remove appendices
Cut down the length some more

In this paper we study a central property of probabilistic programs: \emph{termination}.
When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
(Yet none of these systems provides any support for the development or verification of AST programs.)

Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
A recent result \cite{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

@@

A possible approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. 
In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

Our first answer is the notion of \emph{sparse ranking function}. 
Most of the individual execution steps of a typical program are trivial and easy to mentally skip over. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function. This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. 
This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. 
Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}

Basic (non-random) lambda calculus has the very useful property of confluence, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, confluence may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination. 
The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

@@

\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendix.

>>>>

In this paper we study a central property of probabilistic programs: \emph{termination}.
When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
(Yet none of these systems provides any support for the development or verification of AST programs.) Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08, DBLP:conf/esop/MakOPW21}).

@@

Our approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. 
Equivalently, the program's behavior is used to define an associated (ranking) \emph{supermartingale}. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

Our first answer is the notion of \emph{sparse ranking function}. 
Most of the individual execution steps of a typical program are trivial and easy to ignore. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. Yet, the existence of a sparce ranking function still implies AST (\cref{thm:partial implies rankable}). This also makes this method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the 1D unbiased random walk. 
This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. 
Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}

Basic (non-random) lambda calculus has the very useful property of confluence, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, confluence may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem (\cref{thm:confluent ranking}), that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. 
The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

@@

\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the complete version of this paper on arXiv.\cite{arxiv}

### 2021-04-29 LICS21/related.tex
Cut down the length some more

The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples. Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine. A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.

>>>>

The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but there are several cases where it is a more restrictive than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, applying the function before reducing its argument would not cause any problematic duplication of random samples. A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.
